---
title: "Dancing with the Swans"
author: "Joel Anderson"
date: "2/23/2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## Black Swans
The term "Black Swan" originates from the belief that all swans were white because those were the only ones observed by westerners.  This belief continued until 1697 when black swans were found in Australia, rewriting accepted zoology in one swipe.  The term becomes a metaphor for unprecedented and unexpected events and eventually the title of a best selling book by Nasim Taleb.  
  
There are several characteristics that all Black Swans have in common. The first is that it comes as a surprise to the observer.  It is an outlier outside the realm of regular expectations because nothing in the past points to its existence.  Typically this is because the time frame of which data is available is too short to contain one of these Black Swans due to their rarity.  
  
The second factor that all Black Swans have in common is that they are rationalized after the fact, making them "predictable" in retrospect in the observers mind.  Explanations are created to convince the observer that if they had only looked, the evidence was there the whole time.  Even though there is nothing in the past that points to it.
  
The third characteristic is that it has extreme impact beyond anything ever observed previously.  The consequences can be up to thousands of times a "typical" incident.  The three ingredients to a Black Swan are the unpredictability,  consequences not seen in any historical record and rationalization after the fact.  
  
## Black Swans and Statistics  
Your traditional statistics (i.e. the normal/Gaussian distribution) serve you well if you never venture far from the average. It can be seen in the following examples that catastrophic risk does not follow Gaussian statistical norms.  The Gaussian distribution is very common in any application where the effects are additive.  So common that it often referred to as the Normal distribution.  This works very well in applications where the deviations from the mean are small.  The classic example of this is people's height.  The average height of men is about 5' 10" with a standard deviation of 4".  It would be very rare to come across someone that is more than 2 or 3 standard deviations from that mean.  We don't have giants or Lilliputians walking around!  But in the field of risk, a Black Swan can be a hundred or a thousand times the mean. This is because the variables that drive risk are multiplicative rather than additive.  This means that the results have the ability to increase exponentially.  That is why it is very dangerous to make assumptions about the data to be "normal" in any thing involves risk.  The "normal" assumption while computationally convenient for process that never venture far from the mean where things are predictable but Black Swans live in the wild, out on the tails of the distribution where traditional statistics would predict the likelihood of something of that magnitude is essentially zero.  The first characteristic of Black Swans is that they are a rare bird, your period of time that you have data for might be too short to have ever seen one.  If your entire data set contains nothing but ordinary "White Swans" it does not prove the lack of existence of Black Swans but a single Black Swan is sufficient to disprove that theory of its non-existence and one Black Swan can put a company out of business in one swipe.
  
It can be said that statistics and probability is a statement about the future based on observations in the past. But risk is not a constant thing.  People constantly alter their behavior based on observed or perceived dangers therefor altering future outcomes.  Donald Rumsfeld, former  Secretary of Defense talked about the three types of "knowns".  The first is the "known knowns", things that you know for certain.  The second is the "known unknowns", these are the things that you know you don't know but since you are aware of them you could be acquired through time, effort or other resources.  The third and most consequential is the "unknown unknowns", these are the things that you don't know you don't know - you are completely unaware of them in any way. This is the origins of Black Swans.  For if the factors that could lead to a Black Swan were known the end user would of made adjustments to their behavior to account for them.  But instead they careen ahead attempting to mitigate and prevent the directly perceived more common risk factors that are seen on a regular basis, in the meantime just out of sight could be an event that end their very existence. This is known as availibility bias, the more recent are given more credibility even if eveidence to the contrary exists.  An example would be people that they say they would never swim in the ocean after hearing of a shark attack, even though it is an extremely rare event when compared to things like being attacked by dogs.

## Catastrophic Risk  
It is well known that casinos spend significant sums of money on surveillance and technology to prevent losses due to cheating at the gambling tables.  Yet the biggest loss to a casino came not from cheating, but from the loss of headline entertainment act when Roy Horn one half of the Siegfried and Roy was attacked on stage by one of the tigers in their act.  The total economic loss is estimated at \$100 million dollars when the show closed permanently with no immediate replacement.  The casinos were building defenses against the most common threat (gambling cheats) but were completely blind to the possibility that the loss of this one act impact the casino for years to come and in ways they hadn't even thought of.  Like most Black Swans it comes from an area that is totally non-existent in the historical data.

While history is not a good predictor of the future, most often that is all is available to mine for data.  In the case of pipeline risk PHMSA publishes all pipeline incident data.  An incident is an accidental release from a pipeline that includes any of the following: Greater than $50,000 in property damage, an injury or fatality.  Then if you were to take that data and scale the injuries and fatalities to dollar values using an appropriate multiplier and come up with a total consequence of failure (CoF), the data would have a histogram as shown in Figure 1.  Note: the range of data for the histogram extends out to 10^9 but for the purposes of scale it is truncated at 10^7.
  
```{r import data, fig.cap="Figure 1: CoF Data Density Curve"}
library(dplyr)
library(ggplot2)
#library(readr)
library(MASS)
library(ggpubr)
library(data.table)
library(car)
#setwd("D:/Documents/Analysis/catastrophic")
#setwd("~/PPIM2019")
incidents <- fread("all_combine.csv")
incidents <- incidents[is.na(Year)==FALSE]
#incidents <- incidents[1:2716,] #Drop 40 NAs at the end

incidents[,inflation_adjstd :=incidents$TOTAL_COST_IN84*2.49] #Per CPI calculator adjusted 1984 to 2019
incidents[, CoF:=inflation_adjstd+INJURE*0.33*10e6+FATAL*10e6]
incidents <- incidents[is.na(CoF)==FALSE]

incidents %>% ggplot(aes(CoF))+geom_histogram(bions=40,fill='#FF3030', col='black', alpha=0.8)+geom_rug(col='blue')+theme_bw()+xlim(0,1e8)+ylim(0,60)+labs(title = "Consequence of Failure", subtitle = "PHMSA Gas Transmission Incidents 1986 - 2019", x="CoF (log Scale)")

```

In the field of risk is that the majority of the data will almost always lie in a predictable range near the lower end of the scale but the incidents that should keep one up at night are the ones to the far right which are much rarer but the consequences can be hundreds or thousands of times larger than the average.  In the CoF data in Figure 1 the median incident is around \$170,000 and the maximum is over \$1 billion a difference of over 6,000 times the median.  This is common in any application of risk, not just pipeline risk.  
  
If a normal distribution was fit to something like the incident CoF data it would have the appearance of being a good fit through the middle of the curve with some outliers at the higher end.  But if you were to magnify this tails area you would see that there are far more of these extreme events than would be expected based on a normal distribution. This why the normal distribution is only suitable model for such events that phenomena based on sums and averages but a really poor model for Black Swans where the effects are multiplied by magnitudes of order between the mean and the extremes. Traditional statistics would predict the likelihood of something of the magnitude of these consequences would be effectively zero yet there are several examples in the 33 years of data. In fact the data count departs from the predicted count by a factor of 10 at the higher CoF values.
  
In the next example data of earthquake insurance loss ratios (claims paid divided by premiums received).  If you were an insurance analyst and were looking at the previous 23 years of data in 1993 and trying to determine the risk to the company for the upcoming year what would you recommend to your management?  The highest ratio seen in the historical data is one year where it exceeded 100 and the next highest is about 50 with the majority less than 10.  A person unaware the nature of catastrophic risk might say that the probable maximum loss would be around 100 and the expected loss less than 20.
  
```{r insurance losses, fig.cap="Figure 2: Insurance Losses"}
#setwd("~/PPIM2019")
insurance <- fread("earthquake_loss.csv")
ggplot(insurance[-24], aes(year, loss_ratio))+geom_bar(stat = "identity", fill='steelblue2',col='black')+theme_bw(16, "serif")+scale_y_continuous(breaks=seq(0,140,by=20))+labs(title = "Earthquake Insurance Loss Ratios by Year",subtitle = "1971 - 1993", x="Year", y="Loss Ratio")
#ggplot(insurance[-24], aes(loss_ratio))+geom_histogram( fill='steelblue2',col='black')+theme_bw()+scale_y_continuous(breaks = seq(0,10, by=2))
```
  
The conclusion of this story is that in 1994 the Northridge Earthquake hit with a loss ratio of 2273, almost 700 times the median loss ratio of the previous twenty-three years.  Even if someone was to assume worst case was the mean plus three standard deviations, which under a normal distribution covers 99.7% of the data they would have still been off by a factor of more than 20.  It can be seen from this insurance example that catastrophic risk does not follow Gaussian statistical norms.  The Gaussian distribution is very common in any application where the effects are additive.  So common that it often referred to as the Normal distribution.  This works very well in applications where the deviations from the mean are small.  The classic example of this is people's height.  The average height of men is about 5' 10" with a standard deviation of 4".  It would be very rare to come across someone that is more than 2 or 3 standard deviations from that mean.  We don't have giants or Lilliputians walking around!  But in the field of risk, a Black Swan can be a hundred or a thousand times the mean. This is because the variables that drive risk are multiplicative rather than additive.  These means that the results have the ability to increase exponentially.  
  
